Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-02-21 paper/main.tex

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    completing three 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about lecture~1, 5
    questions about lecture~2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (lectures~1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(lecture~1, lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., lecture~2).

@@

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture~1 are strongly correlated with the average topic weights
from lecture~1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from lecture~2 are
strongly correlated with the average topic weights from lecture~2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with 
their non-matching question sets (lecture~1 video vs.~lecture~2 questions: $r(13) = 
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; lecture~2 video vs.~lecture~1 questions: $r(13) 
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the lecture~1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the lecture~2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., lecture~1 video vs.~lecture~2 questions; lecture~2 video
vs.~lecture~1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces} (i.e., lecture~1),
15 about the conceptual content of \textit{Birth of Stars} (i.e., lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions and answer choices may be found in Table~\questions.

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing lecture~1, the second between
lectures~1 and 2, and the third after viewing lecture~2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about lecture~1, 5 questions about
lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.

>>>>

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    completing three 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about Lecture~1, 5
    questions about Lecture~2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overlap
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (Lectures~1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(Lecture~1, Lecture~2, and general physics knowledge) on each of the three
quizzes. Quiz~1 was intended to assess participants' ``baseline'' knowledge
before training, Quiz~2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., Lecture~1), and Quiz~3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., Lecture~2).

@@

We hypothesized that a topic model trained on transcripts of the two lectures
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices), then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question set (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from Lecture~1 are strongly correlated with the average topic weights
from Lecture~1 questions ($r(13) = 0.809,~p < 0.001$, 95\% confidence interval
(CI)~$= [0.633,~0.962]$), and the average topic weights from Lecture~2 are
strongly correlated with the average topic weights from Lecture~2 questions
($r(13) = 0.728,~p = 0.002$, 95\% CI~$= [0.456,~0.920]$). At the same time, the
average topic weights from the two lectures are \textit{negatively} correlated with 
their non-matching question sets (Lecture~1 video vs.~Lecture~2 questions: $r(13) = 
-0.547,~p = 0.035$, 95\% CI~$= [-0.812, ~-0.231]$; Lecture~2 video vs.~Lecture~1 questions: $r(13) 
= -0.612,~p = 0.015$, 95\% CI~$= [-0.874,~-0.281]$), indicating that the topic model
also exhibits some degree of specificity. The full set of pairwise comparisons
between average topic weights for the lectures and question sets is reported in
Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or question set) reflects about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variances in topic weights (across time or
questions) between the lectures and questions. The variability in topic
expression (over time and across questions) was similar for the Lecture~1 video
and questions ($r(13) = 0.824,~p<0.001$, 95\% CI~$= [0.696,~0.973]$) and the Lecture~2
video and questions ($r(13) = 0.801,~p<0.001$, 95\% CI~$= [0.539,~0.958]$). 
However, as reported in Figure~\ref{fig:topics}B, the variability in
topic expressions across \textit{different} videos and lecture-specific
questions (i.e., Lecture~1 video vs.~Lecture~2 questions; Lecture~2 video
vs.~Lecture~1 questions) were negatively correlated, and neither
video's topic variability was reliably correlated with the topic variability
across general physics knowledge questions. Taken together, the analyses
reported in Figures~\ref{fig:topics} and~\topicWeights~indicate that a topic
model fit to the videos' transcripts can also reveal correspondences (at a
coarse scale) between the lectures and questions.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces} (i.e., Lecture~1),
15 about the conceptual content of \textit{Birth of Stars} (i.e., Lecture~2), and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions and answer choices may be found in Table~\questions.

Over the course of the experiment, participants completed three 13-question
multiple-choice quizzes: the first before viewing Lecture~1, the second between
Lectures~1 and 2, and the third after viewing Lecture~2
(Fig.~\ref{fig:experiment}). The questions appearing on each quiz, for each
participant, were randomly chosen from the full set of 39, with the constraints
that (a) each quiz contain 5 questions about Lecture~1, 5 questions about
Lecture~2, and 3 questions about general physics knowledge, and (b) each
question appear exactly once for each participant. The orders of questions on
each quiz, and the orders of answer options for each question, were also
randomized. Our experimental protocol was approved by the Committee for the
Protection of Human Subjects at Dartmouth College. We used the experiment to
develop and test our computational framework for estimating knowledge and
learning.
